### 2014-06-26 src/docs/talks/paris-dauphine-2014/stan-overview.tex
final talk used at Dauphine

\begin{itemize}
\item Stan is an \myemph{imperative} probabilistic programming language
\vspace*{-12pt}
\begin{itemize}\footnotesize
\item  cf., BUGS: declarative; \ Church: functional; \ Figaro: object-oriented
\end{itemize}
\item Stan \myemph{program}
\vspace*{-4pt}
\begin{itemize}\small
\item declares variables
\item codes log posterior (penalized likelihood)
\end{itemize}
\item Stan \myemph{inference}
\vspace*{-4pt}
\begin{itemize}\small
\item MCMC for full Bayesian inference
\item MLE for point estimation
\end{itemize}
\item Stan is \myemph{open source} \ {\footnotesize (BSD core C++, GPLv3 interfaces)}
\\
{\footnotesize hosted on GitHub; uses Eigen matrix lib, Boost, googletest
  C++ lib}
\end{itemize}

@@

\sld{Books and Model Sets}
%
\begin{itemize}
\item 400+ page core language tutorial manual
\item separate installation, getting started, interface manuals 
\\ {\footnotesize  RStan, PyStan, CmdStan}
\item BUGS and JAGS examples (all 3 volumes), 
\item Gelman and Hill, {\slshape Data Analysis Using Regression and
    Multilevel/Hierarchical Models}
\item Wagenmakers and Lee, {\slshape Bayesian Cognitive Modeling}
\item two other books in progress
\end{itemize}

@@

\sld{Interface Languages}

@@

\begin{itemize}
\item \myemph{\tt\bfseries transformed data} (once after data)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare and define transformed data variables
\item {\slshape execute}: execute definition statements, validate constraints
\end{itemize}
%
\item \myemph{\tt\bfseries transformed parameters} (every log prob eval)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare and define transformed parameter vars
\item {\slshape execute}: execute definition statements, validate constraints
\end{itemize}
%
\item \myemph{\tt\bfseries generated quantities} (once per draw, 
  \code{double} type)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare and define generated quantity
  variables
\\
{\footnotesize (for posterior predictions, event probabilities,
  decision making)}
\item {\slshape execute}: execute definition statements, validate constraints
\end{itemize}
%
\end{itemize}

@@

\begin{itemize}
\item \myemph{\tt\bfseries functions} \ (compiled with model)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare and define functions
\item {\slshape execute}: compile with model
\end{itemize}
\vspace*{12pt}
\item Example
\\[6pt]
\begin{minipage}[t]{0.8\textwidth}
\footnotesize
\begin{Verbatim}
functions {

  real relative_difference(real x, real y) {
    return 2 * fabs(x - y) / (fabs(x) + fabs(y));
  }

@@

\sld{Variable and Expression Types}
\\[3pt]
\hspace*{17pt}Variables and expressions are \myemph{strongly, statically typed}.
\begin{itemize}
\item \myemph{Primitive}: {\tt\small int}, \ {\tt\small real}
\item \myemph{Matrix}: {\tt\small matrix[M,N]}, \ {\tt\small vector[N]}, \ {\tt\small row\_vector[M]}
\item \myemph{Bounded}: primitive or matrix, with 
\\ {\tt\small <lower=L>}, \ {\tt\small <upper=U>}, \ {\tt\small <lower=L,upper=U>}
\item \myemph{Constrained Vectors}: {\tt\small simplex[K]}, \ {\tt\small
    ordered[N]},
\\ {\tt\small positive\_ordered[N]}, \ {\tt\small unit\_length[N]}
\item \myemph{Constrained Matrices}: {\tt\small cov\_matrix[K]}, \ {\tt\small
    corr\_matrix[K]}, \ {\tt\small cholesky\_factor\_cov[M,N]}, \
  {\tt\small cholesky\_factor\_corr[K]}
\item \myemph{Arrays:}  of any type (and dimensionality)
\end{itemize}

@@

\begin{itemize}
\item All built-in \myemph{C++ functions}
\\
{\footnotesize C math, TR1, C++11, including all trig, pow, and
  special log1m, etc.}
\item Extensive library of \myemph{statistical functions}
\\
{\footnotesize e.g., softmax,
log gamma and digamma functions, beta functions, Bessel functions}
%
\item Efficient, arithmetically stable \myemph{compound functions}
\\
{\footnotesize e.g., multiply log, log sum of
exponentials, log inverse logit}
\end{itemize}

@@

\begin{itemize}
\item \myemph{Basic arithmetic}: all arithmetic operators
\item \myemph{Elementwise arithmetic}: vectorized operations
\item \myemph{Solvers}: matrix left and right division, (log) determinant,
inverse 
\item \myemph{Decompositions}: QR, Eigenvalues and Eigenvectors, 
\\
Cholesky factorization, singular value decomposition
\item \myemph{Compound Operations}: quadratic forms, variance scaling
\vfill
\item \myemph{Ordering, Slicing, Broadcasting}: sort, rank, block, rep
\item \myemph{Reductions}: sum, product, norms
\end{itemize}

@@

\begin{itemize}
\item Adaptive \myemph{Hamiltonian Monte Carlo} (HMC)
%
\item Adaptation \myemph{during warmup}
\vspace*{-4pt}
\begin{itemize}\small
\item step size adapted to target Metropolis acceptance rate
\item mass matrix estimated with regularization
{\footnotesize
\\ sample covariance of second half of warmup iterations
\\ (assumes constant posterior curvature)
}
\end{itemize}
%
\item Adaptation \myemph{during sampling}
\vspace*{-4pt}
\begin{itemize}\small
\item number of steps
\\
{\footnotesize aka no-U-turn sampler; aka NUTS}
\end{itemize}
%
\item \myemph{Initialization} user-specified or random unconstrained
\end{itemize}

@@

\begin{itemize}
\item Generated quantities block for predictions, decisions,
  and event probabilities
\item Extractors for samples in RStan and PyStan
\item Coda-like posterior summary
\vspace*{-4pt}
\begin{itemize}\small
\item posterior mean w.\ standard error, standard deviation, quantiles
\item split-$\hat{R}$ multi-chain convergence diagnostic (Gelman and ...)
\item multi-chain effective sample size estimation (FFT algorithm)
\end{itemize}
\item Model comparison with WAIC
\\
{\footnotesize (internal log likelihoods; external cross-sample statistics)}
\end{itemize}

@@

\begin{itemize}
\item Posterior mode finding via BFGS optimization
\\ {\footnotesize (uses model gradient, efficiently approximates Hessian)}
\item Disables Jacobians adjusting for parameter transforms
\item Models, data, initialization as in MCMC
\vfill
\item Near Future:
\vspace*{-4pt}
\begin{itemize}\small
\item  Standard errors on unconstrained scale
\\
{\footnotesize  (estimated using curvature of penalized log likelihood function}
\item Standard errors on constrained scale)
\\
{\footnotesize  (sample unconstrained approximation and inverse transform)}
\item L-BFGS optimizer
\end{itemize}      
\end{itemize}

@@

\begin{itemize}
\item Stan can be used to explore statistical algorithms
\item Models transformed to have support on $\mathbb{R}^n$
\item Once a model is compiled, have
\vspace*{-4pt}
\begin{itemize}\small
\item log probability, gradient, and Hessian
\item data I/O and parameter initialization
\item model provides variable names and dimensionalities
\item transforms to and from constrained representation 
  \\ {\footnotesize (with or without Jacobian)}
\end{itemize}
\item Future:
\vspace*{-4pt}
\begin{itemize}\small
\item second- and higher-order derivatives via auto diff
\end{itemize}
\end{itemize}

@@

\sld{Standard HMC}
\begin{itemize}
\item \myemph{Initialize parameters} diffusely
\\ {\footnotesize Stan's default: \ $q \sim \mbox{\sf
    Uniform}(-2,2)$ on unconstrained scale}
\item For each draw
\vspace*{-3pt}
\begin{itemize}\small
\item leapfrog integrator \myemph{generates proposal}
\item \myemph{Metropolis accept} step ensures detailed balance
\end{itemize}
\vfill
\item \myemph{Balancing act}: small $\epsilon$ has low error, requires many steps
\item Results \myemph{highly sensitive} to step size $\epsilon$ and mass matrix
  $M$
\end{itemize}

@@

\begin{itemize}
\item \myemph{Euclidean} HMC uses \myemph{global mass matrix} $M$
\item Works for densities with \myemph{position-independent curvature}
\item \myemph{Counterexample}: hierarchical model
\vspace*{-4pt}
\begin{itemize}\small
\item hierarchical variance parameter controls lower-level scale
\item mitigate by reducing target acceptance rate
\end{itemize}
\vfill
\item \myemph{Riemannian-manifold} HMC (coming soon)
\vspace*{-4pt}
\begin{itemize}\small
\item automatically adapts to varying curvature
\item no need to estimate mass matrix
\item need to regularize Hessian-based curvature estimate
\\ {\footnotesize (Betancourt arXiv; SoftAbs metric)}
\end{itemize}
\end{itemize}

@@

\begin{itemize}
\item No-U-turn sampler (\myemph{NUTS})
\item Subtle algorithm to maintain \myemph{detailed balance}
\item Move randomly \myemph{forward or backward in time}
\item Double number of leapfrog steps each move (\myemph{binary tree})
\item Stop when a subtree makes a \myemph{U-turn}
\\ {\footnotesize (rare: throw away second half if not end to end U-turn)}
\item \myemph{Slice sample} points along last segment of path
\item \myemph{Generalized} to Riemannian-manifold HMC
\\ {\footnotesize (Betancourt, arXiv)}
\end{itemize}

\sld{Reverse-Mode Auto Diff}
\begin{itemize}
\item Eval gradient in small multiple of function eval time
\\
{\footnotesize (independent of dimensionality)}
\item Templated \myemph{C++ overload} for all functions
\item Code \myemph{partial derivatives} for basic operations
\item Function evaluation builds up \myemph{expression tree}
\item Dynamic program propagates \myemph{chain rule} in reverse pass
\item \myemph{Object-oriented} custom partial propagation
\item Arena-based \myemph{memory management}
\\ {\footnotesize (customize \code{operator new})}
\end{itemize}

\sld{Forward-Mode Auto Diff}
\begin{itemize}
\item Templated \myemph{C++ overload} for all functions
\item Code \myemph{partial derivatives} for basic operations
\item Function evaluation propagates chain rule forward
\item Nest reverse-mode in forward for higher-order
\vfill
\item Faster autodiff rewrite coming in six months to one year
\end{itemize}

@@

\begin{itemize}
\item Fully encapsulates autodiff in C++
\item Autodiff operations are functionals {\footnotesize (higher-order functions)}
\begin{itemize}\small
\item gradients, Jacobians, gradient-vector product
\item directional derivative
\item Hessian-vector product
\item Hessian
\item gradient of trace of matrix-Hessian product
\\ {\footnotesize (for SoftAbs RHMC)}
\end{itemize}
\item Functions to differentiate coded as functors (or pointers)
\\ {\footnotesize (use bind or lambda)}
\end{itemize}

@@

\sld{Parsing and Compilation}
\begin{itemize}
\item Stan code \myemph{parsed} to abstract syntax tree (AST)
\\ {\footnotesize (Boost Spirit Qi, recursive descent, lazy semantic
  actions)}
\item C++ model class \myemph{code generation} from AST
\\ {\footnotesize (Boost Variant)}
\item C++ code compilation
\item Dynamic linking for RStan, PyStan
\end{itemize}

\sld{Coding Probability Functions}
\begin{itemize}
\item \myemph{Vectorized} to allow scalar or container arguments
\\ {\footnotesize (containers all same shape; scalars broadcast as necessary)}
\item Avoid \myemph{repeated computations}, e.g. $\log \sigma$ in
\hspace*{-18pt}
{\small
\begin{eqnarray*}
\textstyle \log \, \mbox{\sf Normal}(y | \mu, \sigma)
& = & \textstyle \sum_{n=1}^N \log \, \mbox{\sf Normal}(y_n | \mu,\sigma)
\\[4pt]
& = & \textstyle \sum_{n=1}^N  - \log \sqrt{2\pi} \ - \log \sigma \ -
\frac{\textstyle y_n - \mu}{\textstyle 2\sigma^2}
\end{eqnarray*}
}
\item \myemph{expression templates} to broadcast and cache scalars
\item \myemph{traits} metaprogram to \myemph{drop constants} (e.g., $-\log
  \sqrt{2 \pi}$) and calculate return types
\end{itemize}

\sld{Models with Discrete Parameters}
\begin{itemize}
\item e.g., simple mixture models, survival models, HMMs,
      discrete measurement error models, missing data
\item \myemph{Marginalize out} discrete parameters
\item Efficient sampling due to \myemph{Rao-Blackwellization}
\item Inference straightforward with expectations
\item \myemph{Difficult} for many users
\end{itemize}

\sld{Models with Missing Data}
\begin{itemize}
\item In principle, missing data just \myemph{additional parameters}
\item In practice, how to declare? 
\vspace*{-3pt}
\begin{itemize}\small
\item observed data as data variables
\item missing data as parameters
\item combine into single vector 
\\ {\footnotesize (in transformed parameters or local in model)}
\end{itemize}
\end{itemize}

@@

\sld{Differential Equation Solver}
\begin{itemize}
\item Auto-diff solutions w.r.t.\ parameters
\item Integrate coupled system for solution with partials
\item Auto-diff coupled Jacobian for stiff systems
\vfill
\item C++ prototype integrated for large PK/PD models
\\ {\footnotesize (project with Novartis for clinical trial)}
\end{itemize}

@@

\sld{Riemannian Manifold HMC}
\begin{itemize}
\item \myemph{SoftAbs} metric
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item Eigendecompose Hessian
\item \myemph{positive definite} with positive eigenvalues
\item \myemph{condition} by narrowing eigenvalue range
\item Betancourt {\slshape arXiv}\ paper
\end{itemize}
\vfill
\item Code complete; awaiting higher-order auto-diff
\end{itemize}

\sld{Thermodynamic Sampler}
\begin{itemize}
\item Physically motivated alternative to ``simulated''
  \myemph{annealing and tempering} (not really simulated!)
\item Supplies external \myemph{heat bath}
\item Operates through \myemph{contact manifold}
\item System relaxes more naturally between energy levels
\vfill
\item Prototypes evaluated
\item Betancourt paper on {\slshape arXiv} (for geometers)
\end{itemize}

>>>>

\begin{itemize}
\item Stan is an \myemph{imperative} probabilistic programming language
\vspace*{-12pt}
\begin{itemize}\footnotesize
\item  cf., BUGS: declarative; \ Church: functional; \ Figaro: object-oriented
\end{itemize}
\item Stan \myemph{program}
\vspace*{-4pt}
\begin{itemize}\small
\item declares variables
\item codes log posterior (penalized likelihood)
\end{itemize}
\item Stan \myemph{inference}
\vspace*{-4pt}
\begin{itemize}\small
\item MCMC for full Bayesian inference
\item MLE for point estimation
\end{itemize}
\item Stan is \myemph{open source} \ {\footnotesize (BSD core C++, GPLv3 interfaces)}
\\
{\footnotesize hosted on GitHub; uses Eigen matrix lib, Boost C++ lib, googletest}
\end{itemize}

@@

\sld{Books and Model Sets}
%
\begin{itemize}
\item 400+ page core language tutorial manual
\item separate installation, getting started, interface manuals 
\\ {\footnotesize  (RStan, PyStan, CmdStan)}
\item BUGS and JAGS examples (all 3 volumes), 
\item Gelman and Hill, {\slshape Data Analysis Using Regression and
    Multilevel/Hierarchical Models}
\item Wagenmakers and Lee, {\slshape Bayesian Cognitive Modeling}
\item two other books in progress
\end{itemize}

\sld{Scaling and Evaluation}
\begin{itemize}
\item Types of scaling
\vspace*{-4pt}
\begin{itemize}\small
\item more data
\item more parameters
\item \myemph{more complex models}
\end{itemize}
\item for MCMC, measure (as function of scaling)
\vspace*{-4pt}
\begin{itemize}\small
\item time to convergence
\item time per effective sample at mixing
\item memory usage
\end{itemize}
\end{itemize}

\sld{Stan vs. BUGS / JAGS}
\begin{itemize}
\item JAGS is the former \myemph{state-of-the-art}, \myemph{black-box} MCMC method
\\ {\footnotesize (Gibbs with slice sampling; chokes w.\ high posterior correlation)}
\item Stan (2.0) vs. BUGS on scaling IRT model
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item Stan about \myemph{10 times faster}
\item Stan memory and time scales linearly
\item linearly scales with number of question answers
\end{itemize}
\vspace*{-3pt}
\vfill
\hspace*{-12pt}
\includegraphics[height=0.6in]{img/irt_scale-eps-converted-to.pdf}
\includegraphics[height=0.45in]{img/irt-memory-2.png}
\item Stan can handle problems that choke BUGS and JAGS
\end{itemize}

@@

\sld{Platforms and Interfaces}

@@

\begin{itemize}
\item \myemph{\tt\bfseries transformed data} (once after data)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare and define transformed data variables
\item {\slshape execute}: execute definition statements, validate constraints
\end{itemize}
%
\item \myemph{\tt\bfseries transformed parameters} (every log prob eval)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare and define transformed parameter vars
\item {\slshape execute}: execute definition statements, validate constraints
\end{itemize}
%
\item \myemph{\tt\bfseries generated quantities} (once per draw, 
  \code{double} type)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare and define generated quantity
  variables; \\
includes pseudo-random number generators
\\
{\footnotesize (for posterior predictions, event probabilities,
  decision making)}
\item {\slshape execute}: execute definition statements, validate constraints
\end{itemize}
%
\end{itemize}

@@

\begin{itemize}
\item \myemph{\tt\bfseries functions} \ (compiled with model)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare and define general (recursive) functions
\\
{\small (use them elsewhere in program)}
\item {\slshape execute}: compile with model
\end{itemize}
\vspace*{6pt}
\item Example
\\[6pt]
\begin{minipage}[t]{0.8\textwidth}
\footnotesize
\begin{Verbatim}
functions {

  real relative_difference(real u, real v) {
    return 2 * fabs(u - v) / (fabs(u) + fabs(v));
  }

@@

\sld{Variable and Expression Types}
\\[3pt]
\hspace*{17pt}Variables and expressions are \myemph{strongly, statically typed}.
\begin{itemize}
\item \myemph{Primitive}: {\tt\small int}, \ {\tt\small real}
\item \myemph{Matrix}: {\tt\small matrix[M,N]}, \ {\tt\small vector[M]}, \ {\tt\small row\_vector[N]}
\item \myemph{Bounded}: primitive or matrix, with 
\\ {\tt\small <lower=L>}, \ {\tt\small <upper=U>}, \ {\tt\small <lower=L,upper=U>}
\item \myemph{Constrained Vectors}: {\tt\small simplex[K]}, \ {\tt\small
    ordered[N]},
\\ {\tt\small positive\_ordered[N]}, \ {\tt\small unit\_length[N]}
\item \myemph{Constrained Matrices}: {\tt\small cov\_matrix[K]}, \ {\tt\small
    corr\_matrix[K]}, \ {\tt\small cholesky\_factor\_cov[M,N]}, \
  {\tt\small cholesky\_factor\_corr[K]}
\item \myemph{Arrays:}  of any type (and dimensionality)
\end{itemize}

@@

\begin{itemize}
\item All built-in \myemph{C++ functions and operators}
\\
{\footnotesize C math, TR1, C++11, including all trig, pow, and
  special log1m, erf, erfc, fma, atan2, etc.}
\item Extensive library of \myemph{statistical functions}
\\
{\footnotesize e.g., softmax,
log gamma and digamma functions, beta functions, Bessel functions of
 and second kind, etc.}
%
\item Efficient, arithmetically stable \myemph{compound functions}
\\
{\footnotesize e.g., multiply log, log sum of
exponentials, log inverse logit}
\end{itemize}

@@

\begin{itemize}
\item \myemph{Basic arithmetic}: all arithmetic operators
\item \myemph{Elementwise arithmetic}: vectorized operations
\item \myemph{Solvers}: matrix division, (log) determinant,
inverse 
\item \myemph{Decompositions}: QR, Eigenvalues and Eigenvectors, 
\\
Cholesky factorization, singular value decomposition
\item \myemph{Compound Operations}: quadratic forms, variance scaling
\item \myemph{Ordering, Slicing, Broadcasting}: sort, rank, block, rep
\item \myemph{Reductions}: sum, product, norms
\item \myemph{Specializations}: triangular, positive-definite, etc.
\end{itemize}

@@

\begin{itemize}
\item Adaptive \myemph{Hamiltonian Monte Carlo} (HMC)
%
\item Adaptation \myemph{during warmup}
\vspace*{-4pt}
\begin{itemize}\small
\item step size adapted to target Metropolis acceptance rate
\item mass matrix estimated with regularization
{\footnotesize
\\ sample covariance of second half of warmup iterations
\\ (assumes constant posterior curvature)
}
\end{itemize}
%
\item Adaptation \myemph{during sampling}
\vspace*{-4pt}
\begin{itemize}\small
\item number of steps
\\
{\footnotesize aka no-U-turn sampler (NUTS)}
\end{itemize}
%
\item \myemph{Initialization} user-specified or random unconstrained
\end{itemize}

@@

\begin{itemize}
\item Generated quantities block for \myemph{inference}
\\ {\footnotesize (predictions, decisions, and event probabilities)}
\item \myemph{Extractors} for samples in RStan and PyStan
\item Coda-like \myemph{posterior summary}
\vspace*{-4pt}
\begin{itemize}\small
\item posterior mean w.\ standard error, standard deviation, quantiles
\item split-$\hat{R}$ multi-chain convergence diagnostic (Gelman and ...)
\item multi-chain effective sample size estimation (FFT algorithm)
\end{itemize}
\item Model comparison with \myemph{WAIC}
\\
{\footnotesize (internal log likelihoods; external cross-sample statistics)}
\end{itemize}

@@

\begin{itemize}
\item Posterior \myemph{mode finding} via BFGS optimization
\\ {\footnotesize (uses model gradient, efficiently approximates Hessian)}
\item \myemph{Disables Jacobians} for parameter inverse transforms
\item Models, data, initialization as in MCMC
\vfill
\item \myemph{Very Near Future}
\vspace*{-4pt}
\begin{itemize}\small
\item  \myemph{Standard errors} on unconstrained scale
\\
{\footnotesize  (estimated using curvature of penalized log likelihood function}
\item Standard errors \myemph{on constrained scale})
\\
{\footnotesize  (sample unconstrained approximation and inverse transform)}
\item L-BFGS optimizer
\end{itemize}      
\end{itemize}

@@

\begin{itemize}
\item Stan can be used to \myemph{explore algorithms}
\item Models transformed to \myemph{unconstrained support} on $\mathbb{R}^n$
\item Once a model is compiled, have
\vspace*{-4pt}
\begin{itemize}\small
\item \myemph{log probability, gradient, and Hessian}
\item data I/O and parameter initialization
\item model provides variable names and dimensionalities
\item transforms to and from constrained representation 
  \\ {\footnotesize (with or without Jacobian)}
\end{itemize}
\item {Very Near Future:}
\vspace*{-4pt}
\begin{itemize}\small
\item second- and \myemph{higher-order derivatives} via auto-diff
\end{itemize}
\end{itemize}

@@

\sld{Standard HMC}
\begin{itemize}
\item \myemph{Initialize parameters} diffusely
\\ {\footnotesize Stan's default: \ $q \sim \mbox{\sf
    Uniform}(-2,2)$ on unconstrained scale}
\item For each draw
\vspace*{-3pt}
\begin{itemize}\small
\item leapfrog integrator \myemph{generates proposal}
\item \myemph{Metropolis accept} step ensures \myemph{detailed balance}
\end{itemize}
\vfill
\item \myemph{Balancing act}: small $\epsilon$ has low error, requires many steps
\item Results \myemph{highly sensitive} to step size $\epsilon$ and mass matrix
  $M$
\end{itemize}

@@

\begin{itemize}
\item \myemph{Euclidean} HMC uses \myemph{global mass matrix} $M$
\item Works for densities with \myemph{position-independent curvature}
\item \myemph{Counterexample}: hierarchical model
\vspace*{-4pt}
\begin{itemize}\small
\item hierarchical variance parameter controls lower-level scale
\item mitigate by reducing target acceptance rate
\end{itemize}
\vfill
\item \myemph{Riemannian-manifold} HMC (coming soon)
\vspace*{-4pt}
\begin{itemize}\small
\item automatically adapts to varying curvature
\item no need to estimate mass matrix
\item need to regularize Hessian-based curvature estimate
\\ {\footnotesize (Betancourt {\slshape arXiv}; SoftAbs metric)}
\end{itemize}
\end{itemize}

@@

\begin{itemize}
\item No-U-turn sampler (\myemph{NUTS})
\item Subtle algorithm to maintain \myemph{detailed balance}
\item Move randomly \myemph{forward or backward in time}
\item Double number of leapfrog steps each move (\myemph{binary tree})
\item Stop when a subtree makes a \myemph{U-turn}
\\ {\footnotesize (rare: throw away second half if not end to end U-turn)}
\item \myemph{Slice sample} points along last branch of tree
\item \myemph{Generalized} to Riemannian-manifold HMC
\\ {\footnotesize (Betancourt, arXiv)}
\end{itemize}

\sld{Reverse-Mode Auto Diff}
\begin{itemize}
\item Eval gradient in small multiple of function eval time
\\
{\footnotesize (independent of dimensionality)}
\item Templated \myemph{C++ overload} for all functions
\item Code \myemph{partial derivatives} for basic operations
\item Function evaluation builds up \myemph{expression tree}
\item Dynamic program propagates \myemph{chain rule} in reverse pass
\item Extensible w.\ \myemph{object-oriented} custom partial propagation
\item Arena-based \myemph{memory management}
\\ {\footnotesize (customize \code{operator new})}
\end{itemize}

\sld{Forward-Mode Auto Diff}
\begin{itemize}
\item Templated \myemph{C++ overload} for all functions
\item Code \myemph{partial derivatives} for basic operations
\item Function evaluation propagates \myemph{chain rule} forward
\item Nest reverse-mode in forward for \myemph{higher-order}
\item \myemph{Jacobians}
\vspace*{-4pt}
\begin{itemize}\small
\item Rerun propagation pass in reverse mode
\item Rerun forward construction with forward mode
\end{itemize}
\vfill
\item Faster autodiff rewrite coming in six months to one year
\end{itemize}

@@

\begin{itemize}
\item Fully encapsulates autodiff in C++
\item Autodiff operations are functionals {\footnotesize (higher-order functions)}
\begin{itemize}\small
\item gradients, Jacobians, gradient-vector product
\item directional derivative
\item Hessian-vector product
\item Hessian
\item gradient of trace of matrix-Hessian product
\\ {\footnotesize (for SoftAbs RHMC)}
\end{itemize}
\item Functions to differentiate coded as functors (or pointers)
\\ {\footnotesize (enables dynamic C++ bind or lambda)}
\end{itemize}

@@

\sld{Parsing and Compilation}
\begin{itemize}
\item Stan code \myemph{parsed} to abstract syntax tree (AST)
\\ {\footnotesize (Boost Spirit Qi, recursive descent, lazy semantic
  actions)}
\item C++ model class \myemph{code generation} from AST
\\ {\footnotesize (Boost Variant)}
\item C++ code \myemph{compilation}
\item \myemph{Dynamic linking} for RStan, PyStan
\end{itemize}

\sld{Coding Probability Functions}
\begin{itemize}
\item \myemph{Vectorized} to allow scalar or container arguments
\\ {\footnotesize (containers all same shape; scalars broadcast as necessary)}
\item Avoid \myemph{repeated computations}, e.g. $\log \sigma$ in
\hspace*{-18pt}
{\small
\begin{eqnarray*}
\textstyle \log \, \mbox{\sf Normal}(y | \mu, \sigma)
& = & \textstyle \sum_{n=1}^N \log \, \mbox{\sf Normal}(y_n | \mu,\sigma)
\\[4pt]
& = & \textstyle \sum_{n=1}^N  - \log \sqrt{2\pi} \ - \log \sigma \ -
\frac{\textstyle y_n - \mu}{\textstyle 2\sigma^2}
\end{eqnarray*}
}
\item recursive \myemph{expression templates} to broadcast and cache scalars,
generalize containers (arrays, matrices, vectors)
\item \myemph{traits} metaprogram to \myemph{drop constants} (e.g., $-\log
  \sqrt{2 \pi}$) and calculate intermediate and return types
\end{itemize}

\sld{Models with Discrete Parameters}
\begin{itemize}
\item e.g., simple mixture models, survival models, HMMs,
      discrete measurement error models, missing data
\item \myemph{Marginalize out} discrete parameters
\item Efficient sampling due to \myemph{Rao-Blackwellization}
\item Inference straightforward with expectations
\vspace*{12pt}
\item Too \myemph{difficult} for many of our users
\\
{\small (exploring encapsulation options)}
\end{itemize}

\sld{Models with Missing Data}
\begin{itemize}
\item In principle, missing data just \myemph{additional parameters}
\item In practice, how to declare? 
\begin{itemize}
\item \myemph{observed} data as data variables
\item \myemph{missing} data as parameters
\item combine into single vector 
\\ {\footnotesize (in transformed parameters or local in model)}
\end{itemize}
\end{itemize}

@@

\sld{Differential Equation Solver}
\begin{itemize}
\item Auto-diff solutions w.r.t.\ parameters
\item Integrate coupled system for solution with partials
\item Auto-diff coupled Jacobian for stiff systems
\vfill
\item C++ prototype integrated for large PK/PD models
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item 
Project with Novartis: 
  longitudinal clinical trial w.\ multiple drugs, dosings, placebo control,
  hierarchical model of patient-level effects, meta-analysis
\item
Collaborators: Frederic Bois, Amy Racine, Sebastian Weber
\end{itemize}
\end{itemize}

@@

\sld{Riemannian Manifold HMC}
\begin{itemize}
\item \myemph{NUTS} generalized to RHMC
\\ {\footnotesize (Betancourt {\slshape arXiv} paper)}
\item \myemph{SoftAbs} metric
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item Eigendecompose Hessian
\item \myemph{positive definite} with positive eigenvalues
\item \myemph{condition} by narrowing eigenvalue range
\item Betancourt {\slshape arXiv}\ paper
\end{itemize}
\vfill
\item Code complete; awaiting higher-order auto-diff
\end{itemize}

\sld{Thermodynamic Sampler}
\begin{itemize}
\item Physically motivated alternative to ``simulated''
  \myemph{annealing and tempering} (not really simulated!)
\item Supplies external \myemph{heat bath}
\item Operates through \myemph{contact manifold}
\item System relaxes more naturally between energy levels
\vfill
\item Prototype complete
\\ {\small (Betancourt paper on {\slshape arXiv}; for geometers)}
\end{itemize}
